---
title: 'Assignment: Spatial Analysis in R'
author: "Sarah Ko"
geometry: margin=2.54cm
output:
  html_document:
    theme: cosmo
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
  always_allow_html: yes
editor_options:
  chunk_output_type: console
---
## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio. 
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
7. After Knitting, please submit the completed exercise (HTML file) to the dropbox in Sakai. Please add your last name into the file name (e.g., "Fay_A09_SpatialAnalysis.pdf") prior to submission.

## DATA WRANGLING

### 1. Prepare the workspace
* Import: tidyverse, sf, and leaflet
```{r Prepare workspace}

# load packages
library(tidyverse)
library(sf)
library(leaflet)
library(mapview)

```

### 2. Read filtered county features into an sf dataframe and plot
In this exercise, we will be exploring stream gage height data in Nebraska, 
as there's been recent floods there. First, we will import from the US Counties  
shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. 
Nebraska's state FIPS code is `31` (as North Carolina's was `37`). 

* Read the cb_2017_us_county_20m.shp shapefile into an sf dataframe
* Filter for Nebraska counties (State FIPS = 31)
* Show the dataset's coordinate reference system
* Plot the records as a map (in any format)
```{r Read the county data into an sf dataframe}
#Read in Counties shapefile into an sf dataframe, filtering for just Nebraska counties

getwd()
# wd should be set to Environmental_Data_Analytics

NE_counties <- st_read('./Data/Spatial/cb_2017_us_county_20m.shp') %>% 
  filter(STATEFP == 31) #Filter for Nebraska counties

#Reveal the CRS of the counties features
st_crs(NE_counties)

#Plot the data

#plot with 'plot'
plot(NE_counties$geometry) # only geometry
plot(NE_counties) # whole dataset

#plot with mapview
mapview(NE_counties)

```
> QUESTION: What is the EPSG code of the Counties dataset? Using http://spatialreference.org, is this a geographic or a projected coordinate system? (Or, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated?  

> ANSWER: The EPSG code of the counties dataset is 4269, this is a geographic coordinate system. This CRS is associated with the NAD83 (i.e. North American Datum 1983) datum. 

### 3. Read in gage locations csv as a dataframe, then display the column names it contains
Next we'll read in some USGS/NWIS gage location data I've added to the `Data/Raw` folder. These 
are in the `NWIS_SiteInfo_NE_RAW.csv` file. (See `NWIS_SiteInfo_NE_RAW.README.txt` for more 
info on this datset.)
* Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe
* Display the column names of this dataset
```{r Read in USGS Gage Locations}
#Read in gage locations csv as a dataframe
NWIS_NE <- read.csv("./Data/Raw/NWIS_SiteInfo_NE_RAW.csv")

#Reveal the names of the columns
names(NWIS_NE)

```
> QUESTION: What columns in the dataset contain the x and y coordinate values, respectively? 

> ANSWER: The column 'dec_long_va' contains the x coordinate values, and the column 'dec_lat_va' contains the y coordinate values. 

### 4. Convert the gage locations dataframe to an sf dataframe of points
* These data use the same coordnate reference system as the counties dataset
* Display the column names of the resulting sf dataframe
```{r}
#Convert to an sf object

NWIS_NE_sf <- st_as_sf(NWIS_NE, coords = c('dec_long_va','dec_lat_va'), crs = st_crs(NE_counties))

#Reveal the structure
dim(NWIS_NE_sf)

#display column names
names(NWIS_NE_sf)

#check the class of the dataframe
class(NWIS_NE)

```
> QUESTION: What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

> ANSWER: The columns 'dec_lat_va' and 'dec_long_va' from the original dataframe have disappeared - they are replaced by a new column 'geometry' in the new sf dataframe.

### 5. Use `ggplot` to plot the gage locations on top of the counties
* Plot the different datasets in different colors
```{r}

ggplot() + 
  geom_sf(data = NE_counties,col='red') + 
  geom_sf(data = NWIS_NE_sf,col='blue') + 
  ggtitle('USGS/NWIS Gage Locations in NE') + 
  theme(plot.title = element_text(hjust = 0.5))

```

### 6. Read in the gage height data and join the site location data to it.  
And finally, we want to attach some gage height data to our site locations. I've constructed a 
csv file listing many of the Nebraska gage sites, by station name and site number along with 
stream gage heights (in meters) recorded during the recent flood event. This file is titled
`NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder. 

* Read this dataset in as a dataframe. 
* Join our site information (already imported above) to these gage height data. 
 * The `site_no` and `station_nm` can both serve as joining attributes. 
 * Construct this join so that the result only includes records where both tables have data. 
* Show the column names in this resulting dataframe
* Once joined, we will again have to convert this product (a dataframe) into a spatial dataframe. Do that.

```{r}
#Read in the data
gage_height <- read.csv("./Data/Raw/NWIS_SiteFlowData_NE_RAW.csv")

#Show the column names
names(gage_height)

#remove na values from dfs
NWIS_NE_remove_na <- NWIS_NE %>% 
  drop_na()
gage_height_remove_na <- gage_height %>%
  drop_na()

#Join location data to it
county_gage_join <- NWIS_NE_remove_na %>%
  inner_join(gage_height_remove_na, by = c("site_no" =  "site_no"), c("station_nm" = "station_nm"))

#show column names in the resulting dataframe
names(county_gage_join)

# check that station names match
test <- county_gage_join$station_nm.x[!(county_gage_join$station_nm.x %in% county_gage_join$station_nm.y)]

# remove column station_nm.y
county_gage_join <- county_gage_join %>%
  select(-station_nm.y)

# rename column station_nm.x to station_nm
county_gage_join = rename(county_gage_join, station_nm = station_nm.x)

#Convert back to sf dataframe
county_gage_sf <- st_as_sf(county_gage_join, coords = c('dec_long_va','dec_lat_va'), crs = st_crs(NE_counties))

```

### 7. Map the pattern of gage height data
Now we can examine where the flooding appears most acute by visualizing gage heights spatially. 
* Plot the gage sites on top of counties
 * Show the magnitude of gage height by color, shape, other visualization technique.

```{r}
#Plot the values

ggplot() + 
  geom_sf(data = NE_counties, col='blue') + 
  geom_sf(data = county_gage_sf, aes(size=gage_ht), alpha=0.5, stroke=0) + 
  ggtitle('USGS/NWIS Gage Heights in NE') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position = "none")

```

## SPATIAL ANALYSIS
Up next we will do some spatial analysis with our data. To prepare for this, we should
transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).

### 8. Transform the counties and gage site datasets to UTM Zone 14N
* Transform each dataset to crs 32614
* Using ggplot, plot the data so that each can be seen as different colors
```{r}
#create sf for gage height

#Transform the counties and gage location datasets to UTM Zone 14
NE_counties_sf_UTM <- st_transform(NE_counties, crs=32614)
gage_height_sf_UTM <- st_transform(county_gage_sf, crs=32614)

#Plot the data
ggplot() + 
  geom_sf(data = NE_counties_sf_UTM,col='red') + 
  geom_sf(data = gage_height_sf_UTM,col='blue') + 
  ggtitle('NE Gage Locations - UTM Zone 14N') + 
  theme(plot.title = element_text(hjust = 0.5))

```
> QUESTION: The shape of Nebraska should look a bit different than the one created in Step 5? Why?

> ANSWER: The map in Step 5 was created with EPSG 4269 - nad83. The map in Step 8 was created with EPSG 32614 - wgs84/utm zone 14n. The different coordinate reference systems change the depiction of the shape of the earth, which causes the shape of the state to look slightly different. The map created with EPSG 32614 appears to sit on a more rounded plane (the latitude and longitude lines are slightly curved) as compared to the map created with EPSG 4269.

### 9. Select the gages falling within a given county
Now let's zoom into a particular county and examine the gages located there. 
* Select Lancaster county from your county sf dataframe
* Select the gage sites falling `within` that county
 * Remember you'll have to create a mask and then apply that mask
* Create a plot showing: 
 * all Nebraska counties, 
 * the selected county, 
 * and the gage sites in that county

```{r}
#Select the county
county_lancaster <- NE_counties_sf_UTM %>%
  filter(NAME == 'Lancaster')

#Select gages within 
#Create a boolean mask 
lancaster_mask <- st_intersects(county_lancaster, 
                            gage_height_sf_UTM,
                            sparse = FALSE)

#Filter the gage dataset using the boolean mask
lancaster_gages <- gage_height_sf_UTM[lancaster_mask,]

#Plot
ggplot() + 
  geom_sf(data = NE_counties_sf_UTM, col='blue') + 
  geom_sf(data = county_lancaster, col='red') +
  geom_sf(data = lancaster_gages, col='dark violet', size=0.7) + 
  ggtitle('Lancaster County Gages') + 
  theme(plot.title = element_text(hjust = 0.5))

```

### 10. Tag each gage site with the name of the county in which it falls
A spatial join (`st_join`) allows us to assign the attributes of an overlapping
feature onto a another feature. We will use to to assign each gage location the 
attributes of the county in which it is located. 
* Spatially join the county features to the gage height features
* Display the list of fields in the resulting dataset
* Map the gage locations, 
  * Include county boundaries
  * Displaying each gage locations county "NAME" as a different color.
  * Display each gage size proportional to its "gage_ht" value
```{r Spatial Join}
#Join features

gage_w_county_data <- st_join(gage_height_sf_UTM, NE_counties_sf_UTM, left = TRUE)

#Show column names
names(gage_w_county_data)

#Plot
ggplot() + 
  geom_sf(data = NE_counties_sf_UTM, col='blue') + 
  geom_sf_label(data = gage_w_county_data, aes(label = NAME, size=2, col=NAME)) +
  geom_sf(data = gage_w_county_data, col='red', aes(size=gage_ht), alpha=0.5, stroke=0) +
  ggtitle('NE Gage and County Data') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position = "none") + 
  ylab("") + xlab ("")

```

### 11. Summarize data by county
Finally, we'll summarize our gage height data by county and then display each county by 
it's mean gage height. 
* Group the spatially joined gage location/county dataset on the county name
 * Compute mean gage height
* Join (non-spatially) this result to our county sf dataframe
 * Prior to joining, you'll need to drop the geometry column from the gage locations
 * To do this, see the `st_drop_geometry` function
* Plot the counties showing mean gage heights for each county
 * Not all counties will have data
```{r}
#Group and summarize

gage_county_grouped <- gage_w_county_data %>%
  filter(!is.na(NAME) & !is.na(gage_ht)) %>% # filter out NAs
  group_by(NAME) %>% 
  summarise(mean_height = mean(gage_ht)) %>% 
  st_set_geometry(NULL) #Convert result to a simple dataframe

#Join summary to County fc
county_mean_gage_join <- NE_counties_sf_UTM %>%
  left_join(gage_county_grouped, by = c("NAME" =  "NAME"))

#Plot
ggplot() + 
  geom_sf(data = county_mean_gage_join, col='blue') + 
  geom_sf_label(data = county_mean_gage_join, aes(label = mean_height)) +
  ggtitle('Mean Gage Height by County') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position = "none") + 
  ylab("") + 
  xlab("")

```

